Methods for the spectroscopic inference of fundamental stellar parameters

### 2014-01-23 paper/paper.tex
Paper updates

\newcommand{\prob}{{\rm prob}}
\newcommand{\qN}{\{q_i\}_{i=1}^N}
\newcommand{\yN}{\{y_i\}_{i=1}^N}
\newcommand{\vt}{\vec{\theta}}
\newcommand{\vg}{\vt_{\star, {\rm grid}}}
\newcommand{\vstar}{\vt_{\star}}
\newcommand{\vN}{\vt_{\rm N}}
\newcommand{\vc}{\vec{c}}
\newcommand{\fM}{f_{\rm M}}
\newcommand{\fD}{f_{\rm D}}
\newcommand{\vD}{\vec{D}}
\newcommand{\dd}{\,{\rm d}}
\newcommand{\trans}{\mathsf{T}}
\newcommand{\Z}{[{\rm Fe}/{\rm H}]}
\newcommand{\A}{[\alpha/{\rm Fe}]}

@@

\title{A method for inference of fundamental stellar parameters using high resolution spectra}
\author{\today{}\\
\medskip
Ian~Czekala\altaffilmark{1} et al.
%Author2\altaffilmark{2},
}

@@

\begin{abstract}
  Matching observed spectra of stars against a grid synthetic spectra is a popular approach to determine fundamental stellar parameters such as effective temperature, surface gravity, and metallicity. However, a fundamental problem with this approach is that the synthetic spectra contain systematic errors such as wrong or missing spectral lines. In the presence of such errors, a simple $\chi^2$ fit between spectra will likely yield biased parameters and incorrect uncertainty estimates. This behaviour is analogous to the way a few strong outliers might severely bias the fit to a straight line. These spectral outliers are difficult to identify \emph{a priori}; procedures such as sigma clipping might be prone to their own systematic effects and discard useful information. We outline an approach to probabilistically identify and account for spectral features that deviate from reality. Such an approach is most needed when operating at high spectral resolution ($R >$ 10,000), since this is when systematic deviations begin to appear most striking, however our technique is still valid for lower resolution spectra as well. We also outline some approaches for spectral fits to brown dwarfs and planets, where instead of spectral lines there may be systematic deviations of a larger swath of the spectrum.
\end{abstract}

%\keywords{}

%\begin{figure}[htb]
%\begin{center}
%\includegraphics{file}
%\caption{}
%\label{fig:}
%\end{center}
%\end{figure}

\section{Introduction}

The fundamental parameters of stars---mass, luminosity, effective photospheric temperature, surface gravity, and metallicity--are of supreme interest to astrophysicists for a wide variety of reasons. A salient example is that most newly discovered exoplanets must necessarily their mass and radius measured relative to the properties of their host stars (e.g. \citealt{tfs+12,blj+12,ssm+13}). The properties of stars are of course also important in their own right because they inform us about our ideas of stellar evolution \citep{dm97, bca+02}. Because stars are the fundamental units of the cosmos, techniques to infer the properties of stars date back to the dawn of astronomy. We briefly review some of the modern and popular techniques in the literature. Approaches can be broadly classified into techniques that compare stars by some broadly defined indices and techniques that attempt to forward-model an observed stellar spectrum by doing a pixel-to-pixel comparison.

\subsection{Similarity techniques}
Many of existing techniques were written to analyze low to medium resolution ($R < $ 10,000) spectra. I am calling them similarity techniques because they compare characteristics of spectra, not actual pixels.

\paragraph{Line indexes} By cataloguing the equivalent widths of chosen spectral lines into indexes, one can correlate the index with stellar properties and identify features that are sensitive temperature, surface gravity, or metallicity. Then, one matches the indecies of a new star against these relationships and derives a set of stellar parameters. A recent example using M dwarfs is \citet{nci+14}. This technique works, but there are many things that could be improved. First, the selection of which lines to measure is arduous and based upon expert knowledge of previously catalogued stars. This is both a disadvantage and an advantage. One one hand by choosing a subset of lines to fit, one may be unwittingly throwing away useful information. On the other hand, it allows the modeller to discard regions of the spectrum that do not follow a simple trend with stellar parameters. Second, while measuring equivalent widths is easy to do, this discards any useful information contained in the shape of the line profile.

\paragraph{Principal component analysis} PCA reduces the dimensionality of the data space and searches for the most salient features of the spectra \citep{ptb14}. How does one properly estimate parameter uncertainties in such an approach? It seems like this approach exists because it is an easy thing to do.

Matrix inversions: MATISSE \citep{rbd06}. This method is similar to PCA in that it derives basis vectors that describe the fundamental parameters of the star. In the PCA analysis, the principal components do not necessarily correspond to true physical quantities. The MATISSE algorithm determines weights that ensure the components are similar to the stellar parameters. They only test the algorithm against input synthetic stellar spectra to detect biases in the parameters, and against the Sun and Arcturus. They do acknowledge that there will be systematic deviations between the synthetic spectra and real spectra, but they do not account for the possible inflation of error estimates.

Given the inherent systematic uncertainties involved in each method, some groups elect to combine the inference from different algorithms, such as (SLOAN SSPP) \citep{lbs+08}. This approach includes machine learning along with many of the previously mentioned techniques.

\subsection{Pixel to pixel techniques using spectral libraries}
Many high quality synthetic spectral grids are now available. Nearly all synthetic stellar libraries are parameterized by the ``fundamental'' stellar parameters of effective temperature, surface gravity, metallicity, and alpha enhancement. We define this combination as 
\begin{equation}
  \vg = \{T_{\rm eff}, \log(g), \Z, \A \}
\end{equation}
Naively, it makes sense to compare an observed spectrum against spectra from the synthetic grid. Using some goodness of fit statistic one can identify the $\vg$ that best reproduce the observed spectrum.

The pixel-to-pixel likelihood function will evaluate the residuals in some form of 
\begin{equation}
  {\rm residual} = \frac{ {\rm data} - {\rm model}}{ {\rm noise}}
\end{equation}
where the best model is defined as one that minimizes the residuals in some sense. We will discuss other goodness of fit functions soon. Each of these quantities are functions of wavelength $\lambda$, so moving forward we will refer to them as the data spectrum $\fD(\lambda)$, the model spectrum $\fM(\lambda | \vt)$, the noise spectrum $\sigma(\lambda)$, and the residual spectrum $R(\lambda | \vt)$.

Most pixel-to-pixel techniques use a $\chi^2$ likelihood function and seek to minimize the square of the residuals
\begin{equation}
  \chi^2 = \sum_\lambda \left [\frac{ \fD(\lambda) - \fM(\lambda | \vstar) }{\sigma(\lambda)} \right ]^2
\end{equation}
It appears that for many techniques, $\sigma(\lambda)$ is the Poisson photon counting errors. For some data reduction pipelines, a per-pixel ``sigma spectrum'' that accounts for spectral extraction errors due to night sky line contamination or low signal to noise is available, and should be used instead. For more discussion of the noise sources beyond statistical Poisson errors, see \S\ref{sec:residuals}.

\paragraph{ULySS} ULySS \citep{kpb+09} is a mature package to make comparisons for individual spectra as well as unresolved stellar populations. Any spectral library can be used as a backend, including empirical spectral libraries such as the one from ELODIE \citep{psk+07}. They explicitly take into account some of the necessary procedures when fitting spectra, such as continuum-multiplication by a polynomial to account for flux calibration errors and rejection of bad data or model points. They use an iterative kappa-sigma clipping method to discard ``bad'' points for the fit. They test that their clipping technique converges to the same answer by starting guesses located at different parameters $\vt$ and seeing if similar lines are masked and similar best fit parameters $\vt$ are obtained.

\paragraph{\citet{mga13}} fit BT-Settl PHOENIX models to M dwarfs. Because high quality, high resolution M dwarf spectra are difficult to obtain owing to their intrinsic faintness, the development of stellar libraries for later spectral types has typically lagged behind. They use a $\chi^2$-like likelihood function and iterate to mask bad pixels. Unlike ULySS, however, they iterate the method against a sample of $\sim 20$ M dwarfs and track which pixels consistently have the high residuals (median residual amount $> 10$\%, or ten times their measurement error) across all stars. Presumably, such a consistent residual would be due to an incorrect physical constant for a certain line or molecular feature. After masking, they repeat the fitting and claim that after two iterations the masks do not change. They note that in the future, rather than selecting good and bad pixels with a binary mask, ``a more nuanced weighting scheme would be to weight each interval (pixel) according to how consistent it is with the models, or test different weighting schemes to see which gives the best agreement with the bolometric temperatures. \textbf{However regions with modest agreement between the real and synthetic spectra may contain more temperature information than regions with slightly better matches, and we have no a priori information about what spectral regions are the most temperature sensitive.}'' Thus, masking ``bad regions'' could very well destroy stellar information in the pursuit of a better ``fit.'' This is a nuanced point that masking schemes generally neglect. We aim to address later in this document.

\paragraph{``Bayesian'' methods} do exist. \citet{sdm+07} developed a hierarchical Bayesian approach that purportedly accounts for both statistical and systematic errors in the estimation of stellar parameters. I need to work through their math more carefully, but it seems as though they use a hierarchical model for the scatter between ``true'' and observed spectrum, and the unknown scatter in the relationship between $\vg$ and the $\fM$. They claim to account for systematics, but it looks like these are systematics errors due to the performance of the instrument, such as a bad flux calibration. They use infrared spectra at low resolution ($R \sim $ 1,200), and a weighted $\chi^2$ function for goodness of fit. As far as I can ascertain, the actually pixel-to-pixel fitting does not account for systematic errors in regions of the model spectrum.

\citet{sb13} present a comprehensive Bayesian method that can fit low or high resolution spectra in addition to photometry and parallax data. They too, mention that they use a $\chi^2$ likelihood. They mention using special masks for the spectra that block noisy and uninformative regions of the spectrum, however they do not go into further detail about how these masks were created or applied. Their unpublished stellar library focuses on parts of the optical spectrum which contain the most stellar information, as determined by experts. As noted by \citet{sb13}, the grids ``sample the wavelength windows around the spectral features important for diagnostic of FGKM stars: 3850 - 4050\AA\ (Ca I lines), 4350 - 4450\AA\ (G-band, CN sensitive), 4600 - 4900\AA\ (H$\beta$), 5100 - 5300\AA\ (Mg I triplet, main gravity diagnostics), 6400 - 6640\AA\ (H$\alpha$), 8400 - 8800\AA\ (Ca II triplet, also used in \emph{Gaia} and in RAVE stellar survey).''

\paragraph{SPC} A very popular method behind the Kepler follow-up program is Stellar Parameter Classification (SPC) \citep{blj+12}). This compares high resolution optical echelle spectra ($R \sim$ 45,000) to high resolution synthetic grids computed with Kurucz models by a cross-correlation method. The best fit parameters and uncertainties are calculated by the relative values of the cross-correlation coefficient. It is uncertain how statistically justified this uncertainty estimate is. Currently SPC uses only a small region of the optical spectrum ($5000 - 5300$\AA) where the spectral library has been carefully tweaked\footnote{changes to line centers and $\log(gf)$ values were on the table} to match high resolution models of the Sun and Vega.

\paragraph{General trends}
Successful pixel-to-pixel methods either mask bad spectral lines or limit themselves to a narrow region of the spectrum, where they can be relatively certain that the spectral library does not contain gross systematical errors. This implicit masking brings up similar concerns to those of \citet{mga13}: how do you know that you are fitting the region of the spectrum that contains the most information about the stellar properties?

\subsection{Line fitting using spectral synthesis}
This is an improvement over the line index method, since now one includes the information contained in the line profile. With quality stellar spectral models, one can identify a few well-modelled lines to fit. By choosing a few good lines to fit, one could determine stellar parameters by matching line profiles for a few select lines \citep{dj03}. However, this is still subject to the need to identify promising regions of the spectrum.

MOOG\footnote{\url{http://www.as.utexas.edu/~chris/moog.html}} \citep{sne73}, SME\footnote{\url{http://www.stsci.edu/~valenti/sme.html}} \citep{vp96}, actually synthesize line profiles on the fly, compared to using a library of spectral models. This has the advantage of enabling individual tweaks to elemental abundances. And interpolation in the model atmosphere space, rather than the spectral line space. However, for many young T-Tauri stars, the spectrum is contaminated by accretion, veiling, and fractional star spot coverage, which can be a nuisance and cause systematic errors if traditional techniques are used. Matching equivalent widths. Interestingly, MOOG sythesizes a spectrum on the fly.

SME allows adjustment of oscillator strengths. We could download the synthesis code for SME and synthesize our own spectra on the fly. The biggest impediment here would be that the simplest atmospheres will not get certain lines correct. The powerful synthesis codes (PHOENIX) will get the atmospheres and lines correct, but they take a long time to run (~hours) per model, and the code is not publicly available. To do a full-spectrum fit for these sorts of codes would likely be too computationally difficult. However, it would allow access to fitting in the parameter space that we want.

\subsection{Stellar parameter estimation in a utopia}

Although the spectral synthesis domain is the correct domain to solve for stellar parameters, the radiation transfer is too slow.

The fitting difficulties of the previous section are a result of trying to fit data to a model in a smaller parameter space. Yes, the combination of $\vg$ is an excellent proxy to describe a star, but in truth (and at high resolution) each star is slightly different. If we think back to first principles, the ``model'' that is actually producing the synthetic spectrum has many, many parameters that have been collapsed down into $\vg$. These parameters include the atmospheric structure (relations of temperature and pressure as a function of stellar depth), individual elemental abundances (all models quote some value relative to solar, a fraction of models allow changes in alpha abundances), and atomic constants describing the opacity contribution of a line. \emph{All} of these parameters have uncertainties in them, some of these are rather dramatic (one continually hears lamentation about the opacities and line lists\dots), yet for any given stellar model these parameters are \emph{fixed}. There may yet be unconsidered stellar physics necessary to explain a small fraction of lines (say for example the cores of Ca lines), but the field of stellar modelling has progressed to the state that I would have the confidence to say that if a group focused all of their attention on a single line, and using only their current physical models and making tweaks to the afore-mentioned lines, they could produce a satisfactory model.

In reality, this has been done using the Kurucz models in an unpublished grid of stellar spectra, over the range $5000 - 5300$ \AA.

The real problem is much more rich, as is the amount of information one could extract from a high resolution optical spectrum. One would want to fit for individual abundances, marginalize over oscillator strengths.

The unfortunate reality is that sampling in the stellar model space for the full optical spectrum is not yet a computational reality. There do exist codes (MOOG) that can do this approach but only for a small ($\sim$ a few lines) at once, and using simplifications. Through collaboration with a stellar atmospheric expert, such an approach may soon be a possibility. If the computation of a given model could be reduced to less than a minute, efficient MCMC sampling parallelized on a cluster (using \texttt{emcee}, for example). However, as it stands without a collaborator, I do not have the expertise nor the time to pursue such an approach. Therefore, we are reduced to fitting in the space of $\vg$ and devising ways to deal with the issues.

\section{General problems with the spectral library approach}

Pixel-to-pixel approaches have the potential to be much more powerful than other historical stellar parameter estimation techniques. When comparing directly in the data space, we have the ability to add in other physics as we see fit.

because we can fit multiple stars at once (say for unresolved binary or multiple systems), fit for velocity components, etc. Many of the most interesting systems for which we would want to determine stellar parameters are in multiple systems anyway, since we can dynamically determine masses for these systems and thus compare to evolutionary models. Determine veiling. And a perhaps wooden-headed belief that using the full spectrum \emph{should} give us a more accurate and less-biased posterior estimate. Such a belief may be foolish and in reality different parts of the spectrum could give completely different parameter estimates.

Many, many methods exist for comparison of stellar spectra, including many ``Bayesian'' approaches in the pixel-to-pixel space. The nature of the Bayesian-ness, however, is usually limited to priors on temperature, surface gravity, and metallicity. These approaches do not explicitly state the likelihood function that they use, or if they do, it is a $\chi^2$ function.

The more successful of these models use low or moderate resolution spectra ($R \sim$ 1,000 to 10,000 ish). Or, a grid over a narrow range of wavelength space with finely tuned atomic constants.

\textbf{The true fundamental problem with this approach is that the stellar spectra are not a perfect representation of reality}. Assumptions about Poisson errors break down, and fits become contaminated by systematics. There are types of errors which occur.

FIGURE: 4 Panel Figure, from left to right, showing Class 0 through III errors. Use Mg b lines, other big line near Mg b triplet, order 24 line, and then Fe I/II line complex that doesn't fit properly.

\paragraph{Class I} A spectral line appears in the data but not in the model spectrum over a wide range of probable stellar parameters (1000 K in $T_{\rm eff}$, and 1 dex in $\log(g)$, $\Z$, and $\A$. Such a type of error is perhaps the easiest to understand and to forgive. For example, such a line could easily appear in the data due to interstellar absorption or telluric absorption, something which a stellar model could never produce. The model could also fail to reproduce a true stellar line due to a missing opacity source.

\paragraph{Class II} A spectral line appears in the model but not in the data over a wide range of probable stellar parameters (1000 K in $T_{\rm eff}$, and 1 dex in $\log(g)$, $\Z$, and $\A$. This is a more puzzling error, because presumably atomic constants would be much more likely to be \emph{missing} than added in by accident. Presumably this is a product of bad atmospheric structure (temperature or pressure is wrong at some point?) that excites a line when it shouldn't otherwise.

\paragraph{Traditional ways to model Class I and Class II errors} are a form of sigma clipping. Mann et al. compares the stars to models over a wide range of parameters. One can track how these lines might be bad because Class I residuals (data - model)/sigma will always be negative (since we are dealing with absorption lines only) and not change significantly with $\vg$, since we the model can only produce continuum. Class II residuals will always be positive and should decrease with decreasing metallicity, since lower metallicity will decrease the strength of all lines.

If a Class I or II error occurs on a line, that region of the spectrum is relatively useless for determining spectral parameters. Authors like Mann et al. or Kholeva et al. find these regions and simply mask them out of the final computation. Kholeva et al., and others approach in an iterative manner. Kholeva et al. approaches are more sophisticated because they track convergence. This is a sigma clipping approach, and while we could go and model these a la Hogg 2010, the approach is not super justified, result is probably not too wrong (similar to actual sigma clipping). Given that Kholeva et al. track convergence maps, I would place a higher degree of confidence in the answer.

>>>>

\newcommand{\prob}{{\rm prob}}
\newcommand{\qN}{\{q_i\}_{i=1}^N}
\newcommand{\qM}{\{q_{im}\}_{i=1,m=0}^{N,M}}
\newcommand{\yN}{\{y_i\}_{i=1}^N}
\newcommand{\vt}{\vec{\theta}}
\newcommand{\vg}{\vt_{\star, {\rm grid}}}
\newcommand{\vpp}{\vt_{\star, {\rm post}}}
\newcommand{\vstar}{\vt_{\star}}
\newcommand{\vN}{\vt_{\rm N}}
\newcommand{\vc}{\vec{c}}
\newcommand{\fM}{ {\bm M}}
\newcommand{\fMi}{M_i}
\newcommand{\fD}{ {\bm D}}
\newcommand{\fDi}{D_i}
\newcommand{\dd}{\,{\rm d}}
\newcommand{\trans}{\mathsf{T}}
\newcommand{\Z}{[{\rm Fe}/{\rm H}]}
\newcommand{\A}{[\alpha/{\rm Fe}]}

@@

\title{Methods for the spectroscopic inference of fundamental stellar parameters}
\author{\today{}\\
\medskip
Ian~Czekala\altaffilmark{1} et al.
%Author2\altaffilmark{2},
}

@@

\section{Introduction}
The fundamental properties of stars---effective photospheric temperature, surface gravity, and metallicity--are crucial for our understanding of pre-main sequence stellar evolution \citep{dm97, bca+02}. The exoplanet community is also intensely focused on deriving accurate stellar properties because the properties of newly discovered exoplanets must necessarily be quoted relative to those of their host stars (e.g. \citealt{tfs+12,blj+12,ssm+13}) and uncertain star properties translate into uncertain planet properties \citep{kan14}. Many techniques\footnote{If you are interested, I have also compiled a more verbose review of modern techniques described in the literature. However, for the sake of brevity I have omitted it from this document.} exist for stellar characterization such as line indexes, machine learning, and spectral synthesis\footnote{Given infinite computational power, we would prefer to fit \emph{all} of the parameters that completely describe a synthetic spectrum: atmospheric structure, individual elemental abundances, and atomic constants describing the opacity contribution of a line. Approximations to the synthesis and efficient MCMC sampling parallelized on a cluster might soon put this approach within reach.}. In order to derive accurate and robust stellar properties for our small sample ($\sim 20$) of T Tauri stars (some binary), we have decided to forward model the observed spectra by drawing from a grid of synthetic spectra.

\begin{table}[!htb]
\begin{tabular}{ll}
\hline
\hline
Symbol & Description\\
\hline
\hline
$i$ & index specifying a pixel\\
$\lambda_i$ & wavelength corresponding to a given pixel $i$\\
$\vg$ & fundamental stellar parameters, $T_{\rm eff}, \log(g), \Z, \A$\\
  & that parameterize a synthetic spectrum from the grid\\
$\vpp$ & stellar parameters $v \sin i$, $v_z$, $A_V$, and $R^2/d^2$ that\\
  & are applied during ``post processing'' of the synthetic spectrum\\
$\vstar$ & $\{\vg,\vpp \}$\\
$c_0, c_1, \ldots$ & Chebyshev polynomial coefficients for residual flux calibration\\
$\vN$ & the set of nuisance parameters composed of $\{c_0, c_1, \ldots, c_N\}$\\
$k(i | \vN)$  & polynomial function that multiplies a synthetic spectrum\\
$\vt$ & the parameters $\{\vg, \vpp, \vN\}$ that completely describe a model spectrum\\
$\fDi$ & data flux for a given pixel, $D(\lambda_i)$\\
$\fD$ & data vector comprised of all $\fDi$, $i = \{0, \ldots, N\}$\\
$\fMi$ & model flux for a given pixel, $M(\lambda_i | \vt)$\\
$\fM$ & model vector comprised of all $\fMi$, $i = \{0, \ldots, N\}$\\
$\sigma_i$ & Poisson noise for a given pixel $i$\\
$R_i$ & residuals $(\fDi - \fMi)/\sigma_i$\\
\hline
\end{tabular}
\caption{Nomenclature used in this document}
\label{tab:nomenclature}
\end{table}

\subsection{Pixel to pixel techniques using synthetic spectral libraries}
\label{sec:pix}
Most forward modeling techniques compare pixels directly using some form of
\begin{equation}
  \chi^2 = \sum_i \left [\frac{ \fDi - \fMi}{\sigma_i} \right ]^2 = \sum_i R_i^2
\end{equation}
where the variables are defined in Table~\ref{tab:nomenclature}. A serious problem with this approach is that synthetic spectra are not perfect: systematic errors invalidate the $\chi^2$ assumption and bias parameter estimates and their uncertainties.

Current methods either circumvent this problem by using high quality synthetic grids that were painstakingly tuned over a narrow wavelength range \citep[$\lesssim 500$ \AA]{blj+12,sb13}, or they use less-perfect grids with full optical coverage and sigma clip the bad regions of the spectrum \citep{kpb+09,mga13}. Masking (either implicitly or explicitly, respectively) is troublesome because, in the words of \citet{mga13}, ``regions with [only] modest agreement between the real and synthetic spectra [that might be masked] may contain more temperature information than regions with slightly better matches, and we have no a priori information about what spectral regions are the most temperature sensitive.'' Masking could very well destroy stellar information in the pursuit of a better fit.

\begin{figure}[!htb]
\begin{center}
\includegraphics{lineclasses}
caption{The four classes of line behavior. \textbf{Class 0}: the model behaves as expected. \textbf{Class I}: a spectral line is present in the data but missing from the model (in this case, the line is also blended). \textbf{Class II}: a spectral line exists in the model where there is only continuum in the data. \textbf{Class III}: lines exist in both the data and model, but some are too strong while others are too weak. The lower panels show the residual spectra in units of the Poisson uncertainty. The data is a spectrum of WASP-14, an F star. The model is from the PHOENIX grid\citep{hwd+13}.}
\label{fig:lineclasses}
\end{center}
\end{figure}

\subsection{Classes of synthetic spectral line behavior}
Illustrated in Figure~\ref{fig:lineclasses} are three ways that synthetic spectra often diverge from reality. When a spectral line appears in the data but not in the model, due to interstellar or telluric absorption or a missing opacity source, we call this a \textbf{Class I} error. When a spectral line appears in the model but not in the data, presumably due to a combination of incorrect atomic constants or atmospheric structure, we label it a \textbf{Class II} error. Both of these types of errors are present over a wide range of probable stellar parameters (1000 K in $T_{\rm eff}$, and 1 dex in $\log(g)$, $\Z$, and $\A$) and there does not exist a $\vg$ that could ever fit these lines, making these regions useless for determining stellar parameters. We know that for a normal AFGK star Class I residuals will always be negative and will not change significantly with $\vg$, while Class II residuals will always be positive and their magnitude should decrease with decreasing metallicity. To help identify Class I and II errors,  we could use the partial derivatives of an individual synthetic spectrum with respect to $\vg$ and a prior on the range of parameters over which the residuals will be correlated to determine which lines show Class I or II behavior.

When some line strengths do not match at a $\vg$ that properly fits most other lines, we term these \textbf{Class III} errors. These types of errors are pernicious because there does exist a $\vg$ that would fit each line individually, but no one $\vg$ can be chosen that fits all the lines. We must then decide which regions to trust and by how much, because Class III regions \emph{still contain information about $\vg$}. In fact, they might contain \emph{the most} information about $\vg$ since these delicate regions are the most sensitive to changes in $\vg$ and are consequently whey they are the hardest to model correctly. In other words, Class III regions might be where the spectrum gradient ($\sum_{j} \partial \fMi/ \partial \vt_{ \star {\rm grid}, j}$) is the largest.

\section{In the limit of a perfect stellar model}
To generate a model spectrum $\fM(\vg, \vpp)$, we use $\vg$ to linearly interpolate a synthetic spectrum from the grid, and then use $\vpp$ to apply instrumental and rotational broadening in the Fourier domain, Doppler shift, correct for extinction, and then downsample to the exact pixels of the data spectrum using spline interpolation\footnote{Because we are dealing with flux-calibrated data and synthetic spectra, we \emph{resample} to lower resolution. We do not \emph{rebin} because we are dealing with flux-density, not photon counts}.

To account for imperfect flux calibration, we extend the approach of \citet{elh+06} and multiply $\fM(\vg, \vpp)$ by $k(\vN)$, which is a Chebyshev polynomial described by a set of $N$ Chebyshev coefficients. We can either sample in the coefficients or analytically marginalize over them. We favor this approach over the traditional technique of normalizing to the stellar continuum due to concerns about correctly placing the pseudo-continuum in M stars. From now on we refer to $\fM(\vt) = k(\vN) \fM(\vg, \vpp)$. The mathematical details are in \S~\ref{sec:chebyshev}.

Currently we are sampling this $\chi^2$ posterior with the MCMC ensemble sampler \citep{gw10} implemented in \texttt{Python} by \citet{fhl+12} as \texttt{emcee}. In this section, we described our approach to sample the posterior assuming perfect synthetic spectra. However, if one does a straight up $\chi^2$ fit to our spectra, one can get posteriors that have 1-sigma contours of $\pm 1$ K and $\pm 0.01$ dex in gravity and metallicity. This would be as precise as stellar parameters obtained with astroseismology! Clearly there is something going wrong here.

\section{Beyond a perfect stellar model}
\label{sec:residuals}
\subsection{Additional sources of noise}
\begin{itemize}
  \item CCD readnoise and imperfect 1D spectrum extraction over night sky lines will almost definitely increase the noise in a pixel beyond theoretical Poisson limits. Normally this information will be encapsulated in a ``sigma spectrum'' output by IRAF, but unfortunately we do not have access to this for our current dataset.
  \item The stellar continuum can flicker/jitter (as shown by Kepler), which would produce a small departure from a perfect continuum. This might explain why even in the Class 0 region of our spectrum (Figure~\ref{fig:lineclasses}, left panel) the residuals are slightly larger than 1 $\sigma$ on average. For our T Tauri stars this effect may be more pronounced.
  \item When we interpolate a synthetic spectrum from the library to a specific $\vg$, we will introduce interpolation errors depending on how close we are to a grid point in the library. \citet{hus12} investigate the possible range of interpolation errors at specific $\vg$ by dropping out a spectrum from the grid, interpolating over it, and then comparing the true spectrum to the interpolated spectra. For hotter stars with well defined continua, we find the maximum error is $\lesssim 5$\%, while for M stars the error could be as large as $\sim 20$\% for some spectral features. By doing this interpolation test across the grid, we could determine an ``interpolation error spectrum'' that would encapsulate the maximum error possible for an interpolated spectrum, weighted by its distance from the nearest grid points. If a typical spectrum has signal/noise of $\sim 30$ per pixel (or noise/signal ratio of $1/30 \approx 3$\%), an interpolation noise contribution of $\sim 5$\% the level of the signal level would be significant. 
\end{itemize}
However, we have tried arbitrarily increasing the Poisson errors by a factor of 2 or 3 and it didn't significantly change the results, suggesting that something more drastic is needed beyond simply adding in more noise.

\subsection{Gallery of likelihood functions}
\label{sec:gal}
A problem with the $\chi^2$ likelihood function is that the Gaussian tails rapidly decline, making outliers extremely improbable. A common approach for ``robust regression'' is to use a likelihood function with fatter, more forgiving, tails. Although not rigorously motivated, this is can be a quick thing to try. We explored several likelihood functions including a combination of Gaussian with exponential wings, a Lorentzian (Cauchy) function, and a function described  by \citet{ss06}, which treats the noise estimate for each data point suspiciously. In Sivia's model, the true noise $\sigma$ can be no smaller than some original estimate $\sigma_0$, but it may be much larger. By putting a prior on $\sigma$, 
\begin{equation}
  p(\sigma_i | {\sigma_0}_i) = \left \{ \begin{array}{cc}
    {\sigma_0}_i/\sigma_i^2 & \sigma_i  \leq {\sigma_0}_i\\
    0                 & \textrm{else}
  \end{array}
    \right.
\end{equation}
and marginalizing over the $\chi^2$ likelihood, the new likelihood function becomes
\begin{equation}
p(\fDi | \fMi, {\sigma_0}_i) = \frac{1}{ {\sigma_0}_i \sqrt{2 \pi}} \left [ \frac{1 - e^{-R_i^2/2}}{R_i^2} \right ] 
\end{equation}
where $R_i = (\fDi - \fMi)/{\sigma_0}_i$. This function has even fatter tails than the exponential model and seems to give results less biased by outliers, however it is not truly satisfying. Which lines contributed most to the posterior? What about the wings of class I and II errors? These individual pixels might still be given a weight of ${\sigma_0}_i$ because they are not nearly as bad as the line centers.

Other functions we might try in the future include a Student-$t$ distribution, a linear combination of Gaussians, or the $\chi^2_Q$ function described in \citet{hbl10}. While ``softer'' likelihood functions do help somewhat, we would like to move beyond tweaking the fitting function towards an approach that properly incorporates our knowledge of the systematic errors of the synthetic grids. In the remaining two sections of this document we outline potential ideas that we might try next, but have not yet implemented in code.

\section{Bad data mixture model}
The bad data model \citep{pre97,hbl10} is a powerful framework for forward modeling data, which allows the modeler to explicitly account for outliers using a generative model. In the simplest mixture model, each pixel has a binary flag $q_i$ which specifies whether that data point is good or bad. Then, for the entire data set we have an ensemble of $\qN$. If this ensemble of $\qN$ is chosen appropriately, and the good pixels are flagged as good and the bad pixels are flagged as bad, the posterior will be maximized. We adopt $P_B$ for the \emph{a priori} probability that a given pixel will be bad, $Y_b$ for the mean flux of the bad pixels, and $V_b$ for their variance. Written out explicitly, this is
\begin{multline}
  {\cal L} = p({\bm D} | \vt,\qN, Y_b,V_b) = \prod_{i=1}^N \left [ \frac{1}{\sqrt{2 \pi} \sigma_i} \exp \left[ - \frac{(\fDi - \fMi)^2}{2 \sigma_i^2} \right] \right ]^{q_i}  \times \\
  \left [ \frac{1}{\sqrt{2 \pi (V_b + \sigma_i^2)}} \exp \left[ - \frac{(\fDi - Y_b)^2}{2 (V_b + \sigma_i^2)} \right]
  \right ]^{1 - q_i}
\end{multline}
If we put a Bernoulli prior on the $q_i$
\begin{equation}
  p(q_i | P_b) = [1-P_b]^{q_i} P_b^{[1-q_i]}
\end{equation}
the probability of any given ensemble of $\qN$ is 
\begin{equation}
  p(\qN | P_b) = \prod_{i=1}^N [1-P_b]^{q_i} P_b^{[1-q_i]}
\end{equation}
We can analytically marginalize over the settings for a single flag by a discrete sum over the two states of $q_i = 1$ and $q_i = 0$. 
\begin{multline}
  p(\vt, P_b, Y_b, V_b | D_i)  =  \int p( \vt, q_i, P_b, Y_b, V_b| D_i) {\rm d} q_i \\
  = p(\vt, q_i=1, P_b, Y_b, V_b | D_i) + p(\vt, q_i=0, P_b, Y_b, V_b | D_i) \\
  = \Bigl[ p_{\rm good}(D_i | \vt)\, p(q_i=1 | P_b) +  p_{\rm bad}(D_i | Y_b, V_b)\, p(q_i=0 | P_b)\Bigr] \times p(\vt ,P_b, Y_b,V_b) 
\end{multline}

\begin{multline}
  p(\vt,P_b,Y_b,V_b|\,{\bm D}) \propto \prod_{i=1}^N \Bigl [ [1 - P_b] p_{\rm good}(D_i |\vt)  + P_b p_{\rm bad}(D_i | Y_b, V_b)\Bigr ] \times p(\vt,P_b, Y_b, V_b)\\
  \boxed{
    = \prod_{i=1}^N \left\{ \frac{1 - P_b}{\sqrt{2 \pi \sigma_i^2}} \exp \left[ - \frac{(\fDi - \fMi)^2}{2 \sigma_i^2} \right] + \frac{P_b}{\sqrt{2 \pi (V_b + \sigma_i^2)}} \exp \left[- \frac{(\fDi - Y_b)^2}{2 (V_b + \sigma_i^2)} \right] \right\}
  \times p(\vt,P_b,Y_b,V_b)}
\end{multline}

In its simplest form, the bad data model results in a likelihood function that is a linear combination of Gaussians. Although this functional form might be similar to some of the likelihood functions in \S~\ref{sec:gal}, the benefit of this bad data approach over a gallery of likelihood functions is that it is motivated by the characteristics of the bad data ($P_B$, $Y_b$, and $V_b$). If we wish to add additional information about the characteristics of the bad data, we can do so.
